---
title: "Alamar_Blue_Be_Mathing"
output: html_document
date: "2023-04-06"
---

#Written by Alvin Kalathungal and Ashish George
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE,cache = TRUE, warning = FALSE, message = FALSE) ## to save intermediate files and make knitting quicker, not show code or warnings
```

## R Markdown

```{r}
#install.packages("dplyr")
library(dplyr)
#install.packages("tidyverse")
library(tidyverse)
#install.packages("openxlsx")
library(openxlsx)
#install.packages("ggplot2")
library(ggplot2)
#install.packages("ggpmisc")
library(ggpmisc)
#install.packages("magrittr")
library(magrittr)
#install.packages("readxl")
library(readxl)
```


only the date of the file needs to inputted and the input and output files are automatically named. Files would need to follow the same naming convention for this to work in the future.


# read and organize data
renames columns, splits data into standard and not, computes log10 values
```{r}

## change the date to reflect metadata date

date_of_file="20230721" ## date of metadata file used to add unique index plots and output files

input_file_name <- paste0("Alamar_Blue_Titer_Metadata_",date_of_file,".xlsx")
output_file_name <- paste0("Alamar_Blue_Titer_Predictions_",date_of_file,".xlsx")

#getwd()


Alamar_Blue_Titer_Metadata <- readxl::read_excel(input_file_name)
# "Alamar_Blue_Titer_Metadata_20230223.xlsx"

# Import metadata sheet & change naming scheme so R can work
data <- Alamar_Blue_Titer_Metadata %>% 
  rename(cell_count = `Cell Count`,sample_name = `Sample Name`, viral_volume = `Viral Volume`, puro_selection = `Puro selection`) %>% mutate(Fluorescence = as.numeric(Fluorescence), cell_count = as.numeric(cell_count)) 


#Grab all the standards to model curve
standard_set <- data %>% filter(grepl("Standard", sample_name)) %>% na.omit()
standard_set <- standard_set %>% mutate(log_fl = log10(Fluorescence)) %>% mutate(log_cell = log10(cell_count))

#grab all the non-standards(Samples)
data_set <- data %>% filter(!grepl("Standard",sample_name))
data_set <- data_set %>% drop_na(Fluorescence) %>% mutate(log_fl = log10(Fluorescence))

# 
target_moi_for_plots <- c(0.1,0.2, 0.3, 0.4, 0.5, 1.0, 1.5, 2.0) 
target_moi_for_report <- c(0.5, 1.0, 1.5, 2.0) 
```
## Plot standard curve to evaluate if the log-log is fit well by a linear model
-- need to add error if fit is not good.
```{r}
#Plot Standard to see if it is linear on log-log
standard_set %>% ggplot(aes(log_fl,log_cell)) + stat_poly_eq() +geom_point() + geom_smooth(method = 'glm')  
ggsave(paste0("std_curve_",date_of_file,".png" ))


#Make Model from standards
linear_model <-  glm(standard_set$log_cell~standard_set$log_fl, data = standard_set) 
# verify if fit is good
# have R^2 value and formula

#Grab the m and b from y=mx + b
intercept_model <- coef(linear_model)[1]
slope_model <- coef(linear_model)[2]

#have IF statements to check for R^2 is less than .8. To stop the program/notify user. 
#potiental outlier
#Have some info on how comfortable we are with the fl reading range downstream
```
## predict cell counts from alamar blue fluorescence based on above

-- add a flag if we are extrapolating the curve far out of the measured range in the standard curve. 

```{r}
# create predicted cell counts 
predicted_set <- data_set %>% mutate(predicted_cell_log = log_fl*slope_model + intercept_model)

#plot log_fl to log_predicted_cell
# predicted_set %>% ggplot(aes(log_fl,predicted_cell_log)) + geom_point()

#convert predicted counts to not be log form
predicted_set <- predicted_set %>% mutate(predicted_cell_count = 10^predicted_cell_log) %>% drop_na(predicted_cell_count)
# head(predicted_set)

#think about flagging / potiential errors in making this curve. If the FL value was out of range, it should be flagged

```


### calculate lambda, the average MOI for each input viral volume

```{r}
# Ask about replicates, NIC exact same values for PURO selection.
# get the ratio for those who survived treatment, versus no treatment and get MOI from it
#*** filter out NIC

predicted_set %>% group_by(sample_name,Dilution, viral_volume, Rep) %>% summarise(ratio = sum(predicted_cell_count[puro_selection == "Yes"], na.rm = TRUE) / sum(predicted_cell_count[puro_selection == "No"],na.rm = TRUE)) %>% mutate(MOI = -log(1 - ratio)) %>% filter(sample_name != "NIC") %>% group_by(sample_name,viral_volume) -> lambda_set


lambda_set <- na.omit(lambda_set)

#Making copies to run each model
linear_model_set <- lambda_set %>% filter(MOI > 0.001)
log_model_set <- lambda_set %>% filter(MOI > 0.001)

head(lambda_set)



## target_moi_for_plots <- c(min(lambda_set$MOI),target_moi_for_plots) ## will predict for the smallest moi as well to conveniently draw lines


```
# predict viral volume for desired MOI

## Linear Model of MOI vs input viral volume

```{r}
# Ploting the linear model via boxplots
linear_model_set %>%  ggplot((aes(x = viral_volume, y= MOI, group = viral_volume))) + facet_wrap(~sample_name) +geom_boxplot()
```



```{r}
#grabbing all the names in the list 
data_split_linear <- split(linear_model_set, linear_model_set$sample_name)
models_linear <- list()
sample_names_linear <- names(data_split_linear)

#creating a model for linear set

for (i in seq_along(data_split_linear)) {
  models_linear[[sample_names_linear[i]]] <- lm(MOI~viral_volume , data = data_split_linear[[i]])
}
```



```{r}
#creating a new data frame with desired MOIs
## making only upto 2.

new_data <- data.frame(moi = target_moi_for_plots)

predictions <- list()


sample_names_linear <- names(data_split_linear)


for (i in seq_along(models_linear)) {
  # Extract the coefficients from the model
  coefs <- coef(models_linear[[i]])
  
  # Calculate the predicted log_volume values
  predictions[[sample_names_linear[i]]] <- (new_data$moi - coefs[1])/ coefs[2] 
}


```


```{r}
sample_names_linear <- names(predictions) 

# Create a data frame with the sample names and the associated predictions
linear_results <- data.frame(sample_names = rep(sample_names_linear, each = nrow(new_data)),
                      moi = rep(new_data$moi, times = length(sample_names_linear)),
                      volume_linear_model = unlist(predictions))
```



## Log-log fit of MOI vs viral volume
```{r}
log_model_set <- log_model_set %>% mutate(log_MOI = log10(MOI)) %>% mutate(log_volume = log10(viral_volume))

log_model_set %>% ggplot((aes(x = log_volume, y= log_MOI, group = viral_volume))) + facet_wrap(~sample_name) +geom_boxplot()

```

```{r}
data_split_log <- split(log_model_set, log_model_set$sample_name)
models_log <- list()

sample_names <- names(data_split_log)


for (i in seq_along(data_split_log)) {
  models_log[[sample_names[i]]] <- lm(I(log_MOI-1*log_volume) ~ 1, data = data_split_log[[i]])
}

```


```{r}
new_data_log <- data.frame(
  log_MOI = log10(target_moi_for_plots),
  moi=target_moi_for_plots)

predictions_log <- list()

sample_names <- names(data_split_log)

for (i in seq_along(models_log)) {
  # Extract the coefficients from the model
  coefs <- coef(models_log[[i]])
  
  # Calculate the predicted log_volume values
  predictions_log[[sample_names[i]]] <- new_data_log$log_MOI - coefs[1]
}

```



```{r}
sample_names <- names(predictions_log)

# Create a data frame with the sample names and the associated predictions
log_results_set <- data.frame(sample_names = rep(sample_names, each = nrow(new_data_log)),
                      log_moi = rep(new_data_log$log_MOI, times = length(sample_names)),
                      moi=rep(new_data_log$moi, times = length(sample_names)),
                      log_volume = unlist(predictions_log))


```



```{r}
#reverting log values to tangible real world values for lab use
log_results <- log_results_set %>% mutate(volume_log_model = 10^(log_volume))#, moi= 10^(log_moi))

#combining linear model and log model dataset into one table
all_results <- left_join(linear_results,log_results, by=c('sample_names','moi')) 

# removing log_moi and log_volume for concise data. Still will exist in final excel sheet
results<- all_results%>% select(-c(log_moi,log_volume)) 

## rename column
results %<>% 
  rename(sample_name = sample_names)

results %<>% 
  dplyr::mutate(percent_discrepancy= 100*abs(volume_log_model-volume_linear_model)/pmin(volume_log_model,volume_linear_model)   ) 

```

## some QC plots
expand prediction to smallest value of input viral volume as well..
plot on log and linear scales.
put points on the lines

```{r ashish_additional_plot}
ggplot() +
  geom_point( data=linear_model_set,aes(x = viral_volume, y =MOI , color="data")) +
  facet_wrap(~sample_name) +
  geom_point(data = results, aes(x =volume_linear_model,  y =moi, color="linear_prediction"),shape=8 )+
  geom_line(data = results, aes(x =volume_linear_model,  y =moi, color="linear_prediction"),shape=8 )+
  geom_point(data = results, aes(x = volume_log_model, y = moi,color="log_prediction"))+
  geom_line(data = results, aes(x = volume_log_model, y =moi, color="log_prediction"))+
  scale_color_manual( values = c("data"="black","linear_prediction" = "blue", "log_prediction" = "red")) 

ggsave(paste0("Preds_Linear_",date_of_file,".png" ))


```

```{r}

ggplot() +
  geom_point(data=linear_model_set, aes(x = viral_volume, y =MOI , color="data")) +
  facet_wrap(~sample_name) +
  geom_point(data = results, aes(x =volume_linear_model,  y =moi, color="linear_prediction"),shape=8 )+
  geom_line(data = results, aes(x =volume_linear_model,  y =moi, color="linear_prediction") )+
  geom_point(data = results, aes(x = volume_log_model, y = moi,color="log_prediction"),shape=8)+
  geom_line(data = results, aes(x = volume_log_model, y =moi, color="log_prediction"))+
  scale_color_manual( values = c("data"="black","linear_prediction" = "blue", "log_prediction" = "red")) +
  scale_x_continuous(trans='log10') +scale_y_continuous(trans='log10') 


ggsave(paste0("Preds_log-log_",date_of_file,".png" ))
# scale_fill_discrete(labels=c('High Program', 'Low Program'))+



```


```{r}

ggplot() +
  geom_point(data=linear_model_set, aes(x = viral_volume, y =MOI , color="data")) +
  facet_wrap(~sample_name) +
  geom_point(data = results, aes(x =volume_linear_model,  y =moi, color="linear_prediction"),shape=8 )+
  geom_line(data = results, aes(x =volume_linear_model,  y =moi, color="linear_prediction") )+
  geom_point(data = results, aes(x = volume_log_model, y = moi,color="log_prediction"),shape=8)+
  geom_line(data = results, aes(x = volume_log_model, y =moi, color="log_prediction"))+
  scale_color_manual( values = c("data"="black","linear_prediction" = "blue", "log_prediction" = "red")) +
  scale_x_continuous(trans='log10') 

ggsave(paste0("Preds_semilog_",date_of_file,".png" ))
```



```{r}

##Exports as a Excel Sheet. Make sure to change naming scheme to match titer data from the correct date
## write.xlsx(list(results= results,all_results = all_results), "titer_data_20230223.xlsx")  

## don't need allresults sheet and don't need results at all moi values predicted.
results_to_report <-  results %>% 
  dplyr::filter(moi %in% target_moi_for_report)

write.xlsx( results_to_report, output_file_name) 


```

